---
title: "Continuous Glucose Monitoring with Freestyle Libre"
author: Richard Sprague
date: "`r Sys.Date()`" 
output: github_document
---

See [Continous Glucose Monitoring: Start Here](http://richardsprague.com/notes/continuous-glucose-monitoring/)

I've been tracking my glucose levels 24 x 7 using a continuous glucose monitor from Abbot Labs called the [Freestyle Libre](https://www.freestylelibre.us/index.html). 

Read (and edit!) my [Continuous Glucose Monitoring Hackers Guide](https://docs.google.com/document/d/11DFx0E-ZQ-r_D1SqXvMtvkDCjx6j7NevrE43WSaKyJE/edit?usp=sharing) for details for how to get started, plus as many resources as I know about other apps and links that you might find useful for beginning your own CGM analysis.

This is a short R script I use for my analysis.

---

First we read the data and put it into a tidy format in two dataframes:

`libre_raw` : the raw output from a Librelink CSV file. You could just read.csv straight from the CSV if you like.

`activity_raw`: your file containing the metadata about whatever you'd like to track. The following script assumes you'll have variables for `Sleep`, `Exercise`, `Food`, and a catch-all called `Event`. 

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(readxl)
libre_raw <- readxl::read_excel(file.path(Sys.getenv("ONEDRIVE"),"Sprague 2019","Health 2019","Librelink.xlsx"))
libre_raw$`Meter Timestamp` <- lubridate::force_tz(libre_raw$`Meter Timestamp`, "US/Pacific")

activity_raw_2019 <- readxl::read_excel(file.path(Sys.getenv("ONEDRIVE"),"Sprague 2019","Health 2019","Rik Activity 2019.xlsx"),
                                   sheet = "2019")
activity_raw_2018 <- readxl::read_excel(file.path(Sys.getenv("ONEDRIVE"),"Sprague 2019","Health 2019","Rik Activity 2019.xlsx"),
                                   sheet = "2018")

activity_raw <- rbind(activity_raw_2018,activity_raw_2019)

```

Now clean up the data and then set up a few other useful variables.

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(ggthemes)

activity_raw$Start <- activity_raw$Start %>% lubridate::parse_date_time(order = "ymd HMS",tz = "US/Pacific")
activity_raw$End <- activity_raw$End %>% lubridate::parse_date_time(order = "ymd HMS", tz = "US/Pacific")


glucose <- libre_raw %>% select(time = "Meter Timestamp", 
                                scan = "Scan Glucose(mg/dL)",
                                hist = "Historic Glucose(mg/dL)",
                                strip = "Strip Glucose(mg/dL)",
                                food = "Notes")

#glucose$time <- readr::parse_datetime(libre_raw$`Meter Timestamp`,locale = locale(tz="US/Pacific"))

glucose$time <- as_datetime(libre_raw$`Meter Timestamp`, tz = "US/Pacific")
# 
glucose$value <- dplyr::if_else(is.na(glucose$scan),glucose$hist,glucose$scan)

# apply correction for faulty 2019-01-08 sensor
#glucose$value <- dplyr::if_else(glucose$time>as_datetime("2019-01-08"),glucose$value+35,glucose$value)

glucose_raw <- glucose

# libre_raw$`Meter Timestamp` %>% lubridate::parse_date_time(order = "ymd HMS",tz = "US/Pacific")
```

Set up a few convenience functions.

```{r, message = FALSE, warning = FALSE}
# a handy ggplot object that draws a band through the "healthy" target zones across the width of any graph:
glucose_target_gg <-   geom_rect(aes(xmin=as.POSIXct(-Inf,  origin = "1970-01-01"),
                xmax=as.POSIXct(Inf,  origin= "1970-01-01"),
                ymin=100,ymax=140),
            alpha = 0.01, fill = "#CCCCCC",
            inherit.aes = FALSE)


# show glucose levels between start and end times
cgm_display <- function(start=lubridate::now()-lubridate::hours(18),
                        end=now(),
                        activity_df=activity_raw,
                        glucose_df=glucose_raw) {
  ggplot(glucose_df ,aes(x=time,y=value)) + geom_line(size=2, color = "red")+ 
  geom_point(stat = "identity", aes(x=time,y=strip), color = "blue")+
  glucose_target_gg + 
  geom_rect(data=activity_df %>% dplyr::filter(Activity == "Sleep") %>%
              select(xmin = Start,xmax = End) %>% cbind(ymin = -Inf, ymax = Inf),
            aes(xmin=xmin,xmax=xmax,ymin=ymin,ymax=ymax),
            fill="red",
            alpha=0.2,
            inherit.aes = FALSE) +
  geom_rect(data=activity_df %>% dplyr::filter(Activity == "Exercise") %>%
              select(xmin = Start,xmax = End),
            aes(xmin=xmin,xmax=xmax,ymin=-Inf,ymax=Inf),
            fill="blue",
            alpha=0.2,
            inherit.aes = FALSE) +
   geom_vline(xintercept = activity_df %>% 
               dplyr::filter(Activity == "Event" & Comment == "awake") %>% select("Start") %>% unlist(),
             color = "green") +
  geom_vline(xintercept = activity_df %>% 
               dplyr::filter(Activity == "Food") %>% select("Start") %>% unlist(),
             color = "yellow")+
  geom_text(data = activity_df %>%
              dplyr::filter(Activity == "Food") %>% select("Start","Comment") ,
            aes(x=Start,y=50, angle=90, hjust = FALSE,  label = Comment),
            size = 6) +
  labs(title = "Glucose (mg/dL)", subtitle = start) +  theme(plot.title = element_text(size=22))+
    scale_x_datetime(limits = c(start,end),
                     date_labels = "%m/%d %H:%M",
                     timezone = "US/Pacific")
  
}

# returns a dataframe giving all glucose values within "timelength" of a specific activity
food_effect <- function( foodlist = c("Oatmeal","Oatmeal w cinnamon"), activity_df = activity_raw, glucose_df = glucose_raw, timelength = lubridate::hours(2)){
  #food_df <- activity_df %>% dplyr::filter(str_detect(str_to_lower(activity_df$Comment),pattern = foodname))
  food_df <- activity_df %>% dplyr::filter(Comment %in% foodlist)
  food_df$Comment <- paste0(food_df$Comment,rownames(food_df))
  food_df_interval <- interval(food_df$Start,food_df$Start + hours(1))
  food_glucose <- glucose_df %>% dplyr::filter(apply(sapply(glucose_df$time,function(x) x %within% food_df_interval),2,any))
 # food_glucose <- glucose_df %>% dplyr::filter(sapply(glucose_df$time,function(x) x %within% food_df_interval))
  f <- cbind(food_glucose[1,],experiment = "test")
  
  a = NULL
  
  for(i in food_df$Start){
    i_time <- as_datetime(i, tz = "US/Pacific")
    # < rbind(i,a)
    g <- glucose_df %>% dplyr::filter(time %within% interval(i_time - minutes(10), i_time + timelength))
    #print(g)
    p = match(as_datetime(i),food_df$Start)
    f <- rbind(f,cbind(g,experiment = food_df$Comment[p]))
  }
  foods_experiment <- f[-1,]
  foods_experiment
}


```



View the last couple days of the dataset:

```{r, message = FALSE, warning = FALSE}

startDate <- now() - days(2) #min(glucose$time)

#cgm_display(startDate,now()-days(6))

cgm_display(startDate,startDate + days(2))
```

Here's just for a single day:

```{r, message = FALSE, warning = FALSE}



#pdf("icecream.pdf", width = 11, height = 8.5)
cgm_display(start = min(glucose_raw$time),min(glucose_raw$time)+hours(24))
#dev.off()

```


The final full day of the dataset:

```{r, warning = FALSE}
cgm_display(start = max(glucose_raw$time)-days(1), end = max(glucose_raw$time))
```


## Food types

Here's how I look when eating specific foods:

```{r, echo = FALSE, warning = FALSE}


theme_set(theme_stata())

 foodlist <- c("Oatmeal (plain)","Oatmeal w/milk","Oatmeal w/butter", "Oatmeal w/cinnamon")
food_effect(foodlist) %>% group_by(experiment) %>% mutate(delta = (time - min(time))/60) %>% ggplot(aes(x=delta,y=value, color = experiment)) + geom_line(size = 2) + scale_x_continuous() + labs(title = "Glucose after eating oatmeal", x = "Minutes", y = "mg/dL")

foodlist <- c("Potato (reheated)","Lamb + potato")
food_effect(foodlist) %>% group_by(experiment) %>% mutate(delta = (time - min(time))/60) %>% ggplot(aes(x=delta,y=value, color = experiment)) + geom_line(size = 2) + scale_x_continuous() + labs(title = "Glucose after eating potatoes", x = "Minutes", y = "mg/dL")
 


## Beer
foodlist <- activity_raw %>% dplyr::filter(str_detect(str_to_lower(activity_raw$Comment),pattern = "beer")) %>% .$Comment
food_effect(foodlist) %>% group_by(experiment) %>% mutate(delta = (time - min(time))/60) %>% ggplot(aes(x=delta,y=value, color = experiment)) + geom_line(size = 2) + scale_x_continuous() + labs(title = "Glucose after drinking beer", x = "Minutes", y = "mg/dL")
# 

## Pasta
foodlist <- c("Pesto pasta","Pasta whole wheat w/tomato",  "Pasta mushroom chicken")
food_effect(foodlist) %>% group_by(experiment) %>% mutate(delta = (time - min(time))/60) %>% ggplot(aes(x=delta,y=value, color = experiment)) + geom_line(size = 2) + scale_x_continuous() + labs(title = "Glucose after eating pasta", x = "Minutes", y = "mg/dL")
# 

## Apples

foodlist <- c("Apple")
food_effect(foodlist, timelength = hours(1)) %>% group_by(experiment) %>% mutate(delta = (time - min(time))/60) %>% ggplot(aes(x=delta,y=value, color = experiment)) + geom_line(size = 2) + scale_x_continuous() + labs(title = "Glucose after eating an apple", x = "Minutes", y = "mg/dL")
# 

## ramen
foodlist <- c("Ramen", "donut")
food_effect(foodlist, timelength = hours(2)) %>% group_by(experiment) %>% mutate(delta = (time - min(time))/60) %>% ggplot(aes(x=delta,y=value, color = experiment)) + geom_line(size = 2) + scale_x_continuous() + labs(title = "Glucose after eating instant ramen", x = "Minutes", y = "mg/dL")
# 

foodlist <- c("Blueberry Kefir", "Blueberry Kefir w Cinnamon")
food_effect(foodlist, timelength = hours(2)) %>% group_by(experiment) %>% mutate(delta = (time - min(time))/60) %>% ggplot(aes(x=delta,y=value, color = experiment)) + geom_line(size = 2) + scale_x_continuous() + labs(title = "Glucose after drinking kefir", x = "Minutes", y = "mg/dL")
# 

from <- activity_raw %>% filter(Comment == "Apple juice") %>% .$Start

food_effect(foodlist = c("Apple juice","Apple juice + donut")) %>% group_by(experiment) %>% mutate(delta = (time - min(time))/60) %>% ggplot(aes(x=delta,y=value, color = experiment)) + geom_line(size = 2) + scale_x_continuous() + labs(title = "Glucose after apple juice", x = "Minutes", y = "mg/dL")

ggplot(data = glucose_raw %>% filter(time > from
                       & time < (from + hours(2))),
       aes(x=time,y=value)) + geom_line(size=2, color = "red") + glucose_target_gg +
  # scale_x_continuous() + 
  labs(title = "Glucose after drinking apple juice", x = "Minutes", y = "mg/dL")

cgm_display(start = activity_raw %>% filter(Comment == "Apple juice") %>% .$Start, 
            end = activity_raw %>% filter(Comment == "Apple juice") %>% .$Start + hours(2))



cgm_display(start = activity_raw %>% filter(Comment == "Ramen") %>% .$Start, 
            end = activity_raw %>% filter(Comment == "Ramen") %>% .$Start + hours(2))

foods_experiment <- food_effect(foodlist="Ice Cream") 
foods_experiment %>% group_by(experiment) %>% mutate(delta = (time - min(time))/60) %>% ggplot(aes(x=delta,y=value, color = experiment)) + geom_line(size = 2) + scale_x_continuous() + labs(title = "Glucose after eating ice cream", x = "Minutes", y = "mg/dL")


cgm_display(start = activity_raw %>% filter(Comment == "Blueberry-banana") %>% pull(Start),
            end = activity_raw %>% filter(Comment == "Blueberry-banana") %>% pull(Start) + hours(2)
) + labs(x = "time", title = "After a smoothie", subtitle = "Glucose (mg/dL)", y = "mg/dL") + scale_x_datetime(limits = c(activity_raw %>% filter(Comment == "Blueberry-banana") %>% pull(Start),
                                                                                                                             activity_raw %>% filter(Comment == "Blueberry-banana") %>% pull(Start) + hours(2)))


foodlist <- activity_raw %>% dplyr::filter(str_detect(str_to_lower(activity_raw$Comment),pattern = "latte")) %>% .$Comment

food_effect(foodlist) %>% group_by(experiment) %>% mutate(delta = (time - min(time))/60) %>% ggplot(aes(x=delta,y=value, color = experiment)) + geom_line(size = 2) + scale_x_continuous() + labs(title = "Glucose after drinking a latte", x = "Minutes", y = "mg/dL")

food_effect(c("CBD")) %>% group_by(experiment) %>% mutate(delta = (time - min(time))/60) %>% ggplot(aes(x=delta,y=value, color = experiment)) + geom_line(size = 2) + scale_x_continuous() + labs(title = "Glucose after CBD", x = "Minutes", y = "mg/dL")

food_effect(foodlist) %>% group_by(experiment) %>% summarize(max(value), min(value), change = max(value)-min(value))  %>% knitr::kable(caption = "Glucose Changes from drinking a latte")



foodlist <- c("Salmon")
food_effect(foodlist, timelength = hours(1)) %>% group_by(experiment) %>% mutate(delta = (time - min(time))/60) %>% ggplot(aes(x=delta,y=value, color = experiment)) + geom_line(size = 2) + scale_x_continuous() + labs(title = "Glucose after eating salmon", x = "Minutes", y = "mg/dL")
# 
```



## Basic Statistics

What is my average glucose level while sleeping?

```{r}
library(lubridate)

options(scipen = 999)

# all sleep intervals, using the lubridate function lubridate::interval
# sleep_intervals <- interval(activity_raw %>% dplyr::filter(Activity == "Sleep") %>% select(Start)
#                             %>% unlist() %>% as_datetime(),
#                             activity_raw %>% dplyr::filter(Activity == "Sleep") %>% select(End) 
#                             %>% unlist() %>% as_datetime())

activity_intervals <- function(activity_raw_df, activity_name){
  interval(activity_raw_df %>% dplyr::filter(Activity == activity_name) %>% select(Start)
                            %>% unlist() %>% as_datetime(),
                            activity_raw_df %>% dplyr::filter(Activity ==activity_name) %>% select(End) 
                            %>% unlist() %>% as_datetime())
}



glucose %>% filter(apply(sapply(glucose$time,
                                function(x) x %within% activity_intervals(activity_raw,"Sleep")),2,any)) %>% select(value) %>% 
  DescTools::Desc(main = "Glucose Values While Sleeping")

glucose %>% filter(apply(sapply(glucose$time,
                                function(x) !(x %within% activity_intervals(activity_raw,"Sleep"))),
                         2,
                         any)) %>% select(value) %>% 
  DescTools::Desc(main = "Glucose Values While Awake")



glucose %>% filter(apply(sapply(glucose$time,
                                function(x) x %within% activity_intervals(activity_raw,"Exercise")),2,any)) %>% select(value) %>% 
  DescTools::Desc(main = "Glucose Values While Exercising")


```



